---
title: "WORKFLOW for Lake Llaviucu aquatic-terrestrial ecological coupling"
author: 'Author: Xavier Benito (xavier.benito.granell@gmail.com)'
date: 'Date: 29/04/2022'
output:
  html_notebook: default
  pdf_document: default
---

## Introduction
This workflow is written as a R Markdown document to show data analyses to reproduce the manuscript [**'Human practices are behind the aquatic and terrestrial decoupling to climate change in tropical Andes'**](https://www.sciencedirect.com/science/article/abs/pii/S0048969722012074?via%3Dihub). This document shows the most relevant code. See R files within the folder "scripts" for a more specific code. Please refer to the full manuscript for full interpretation and data and plots.

## Study area and datasets description
+ **Lake Llaviucu**: morrained-dammed tropical Lake located in Ecuador (Cajas National Park), at 3115 m of elevation.
Two sediment cores: 
  + **Diatom**: 3-m core covering the last ca 3,000 years, and sampled at 1-cm intervals for diatom analysis [Benito et al. 2021 L&O] (https://aslopubs.onlinelibrary.wiley.com/doi/full/10.1002/lno.11747) 
  + **Pollen**: 11.9-m core covering the last 14,000 years, and sampled at 1-cm intervals for pollen, charcoal and Sporiormella analyses [Nascimiento et al. 2020 QSR] (https://www.sciencedirect.com/science/article/abs/pii/S0277379120304339?via%3Dihub)

## Load packages
```{r packages, echo=FALSE}
library(analogue)
library(tidyverse)
library(vegan)
library(analogue)
library(mgcv)
library(ggplot2)
library(gratia)
```


## Step 1: Estimate Principal Curves (PrC) of the diatom record.
PrC is a nonlinear ordination technique that extracts a single gradient of variation from multivariate data.
```{r principal curves diatoms, echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
#read diatom core datasets
mergedCores <- read.csv("data/mergedCores_counts4.csv") #this is a dataframe with absolute counts 

agedepth <- mergedCores[, names(mergedCores) %in% c("depth", "upper_age", "lower_age", "lake")]
diat <- mergedCores[, !names(mergedCores) %in% c("depth", "upper_age", "lower_age", "lake")]
diat[is.na(diat)] <- 0

diatoms_save <- cbind(agedepth, diat)

changes <- read.csv("data/old_new_nms_cores_counts.csv", stringsAsFactors = FALSE)
#new1: ecological groups
#new2: harmonized taxonomic names

#this is to transform to tidy format, calculate % and subset more common species
new <- diatoms_save %>% 
  gather(key = taxa, value = count, -depth, -upper_age, -lower_age, -lake) %>%
  mutate(taxa = plyr::mapvalues(taxa, from = changes[,1], to = changes$new_2)) %>%
  group_by(depth, taxa, lake, upper_age, lower_age) %>%
  summarise(count = sum(count)) %>%
  filter(!count == "0" ) %>% #this is to remove empty samples (rows)
  filter(!upper_age==0.0) %>% #this is to drop extraneous ages
  ungroup() %>%
  group_by(depth, lake) %>%
  mutate(relative_abundance_percent = count / sum(count) * 100) %>%
  ungroup()

# filter more abundant taxa; format need to be on long-wide format-->no spreaded 
core_common_taxa <- new %>%
  group_by(taxa) %>%
  summarise(max_rel_abund = max(relative_abundance_percent)) %>%
  filter(max_rel_abund >= 5) %>%
  arrange(max_rel_abund) %>%
  pull(taxa)

# select from initial table
core_counts_common <- new %>%
  filter(taxa %in% core_common_taxa) %>%
  mutate(taxa = factor(taxa, levels = core_common_taxa)) %>%
  arrange(taxa)

#make it wide
core_counts_wide <- core_counts_common %>%
  select(depth, lake, upper_age, lower_age, taxa, relative_abundance_percent) %>%
  spread(key = taxa, value = relative_abundance_percent) 

## split cores by lakes and reassemble
coresList <- split(core_counts_wide, core_counts_wide$lake)

# Extract LLaviucu diatom dataset
llaviucu_diatoms <- as.data.frame(coresList[["llaviucu"]])
llaviucu_diatoms[is.na(llaviucu_diatoms)] <- 0

## extract agedepth variables
agedepth <- llaviucu_diatoms[,names(llaviucu_diatoms) %in% c("depth", "lake", "upper_age", "lower_age")]
llaviucu_diatoms <- llaviucu_diatoms[,!names(llaviucu_diatoms) %in% c("depth", "lake", "upper_age", "lower_age")]

#Select most abundant and common species
abund <- apply(llaviucu_diatoms, 2, max)
n.occur <- apply(llaviucu_diatoms>0, 2, sum)
llaviucu_diatoms <- llaviucu_diatoms[, abund>1 & n.occur >2] #more than 1% of RA and present in >2 samples

# Combine dataframe with ages and depths
llaviucu_diatoms <- cbind(agedepth, llaviucu_diatoms)

# Save Diatom data RA
#write.csv(llaviucu_diatoms, "data/diatomsRA.csv")

# Function to fit principal curves
fitPcurve <- function(i, cores, axis, method, ...) {
  axis <- axis[i]
  core <- cores[[i]]
  method <- method[i]
  core <- core[ , -which(names(core) %in% c("depth","upper_age", "lower_age", "lake"))] # drop year & depths vars
  core[is.na(core)] <- 0
  core <- core[, colSums(core) > 0] #select only present species
  core <- decostand(core, method="hellinger") #Hellinger transform relative abundance data
  prc <- prcurve(core, vary = TRUE, trace = TRUE, method = method,
                 axis = axis)
  prc
}

## apply fitCurve  function to each core
axis <- c(1, 1, 1, 1, 1, 1, 1, 1) 
method <- c(rep("ca", 8)) 

corePRC <- lapply(seq_along(coresList), fitPcurve,
                  cores = coresList, axis = axis, method = method)
names(corePRC) <- names(coresList)

# extract data from list
scrs <- lapply(corePRC, scores)
scrs <- lapply(scrs, `[`, , 1)
len <- lapply(scrs, length)
nams <- names(scrs)
depths <- lapply(coresList, `[`, , "depth")
age <- lapply(coresList, `[`, ,  "upper_age")
upper_age <- lapply(coresList, `[`, ,  "upper_age")
lower_age <- lapply(coresList, `[`, ,  "lower_age")
df <- data.frame(Lake = rep(nams, times = len),
                 PrC  = unlist(scrs),
                 Depth = unlist(depths),
                 Age = unlist(age),
                 negAge = - unlist(age),
                 upper_age = unlist(upper_age),
                 lower_age = unlist(lower_age))
rownames(df) <- NULL

#Plot Llaviucu diatom PrC
cores.plot <- ggplot(df[df$Lake=="llaviucu",], aes(x = Age, y = PrC)) +
  geom_point() + geom_line() +
  #coord_flip() +
  facet_wrap(Lake~., scales = "free") +
  ggtitle("Diatoms PrC")
cores.plot

## write this out for use in the GAM modelling script
#saveRDS(df, file = "outputs/PrC-cores-diatoms.rds") #this is including all spp

```

## Step 2: Estimate PrC of the pollen record (including agropatoralism indicators)
```{r principal curves, echo=TRUE, error=FALSE, message=FALSE, warning=FALSE, size="small"}

# read Pollen data (drop Lycopodium counts)
pollen <- read.csv("data/llaviucu_pollen_counts.csv") %>%
  select(everything(), -contains("Lycopodium"))

#read in Charcoal data
charcoal <- read.csv("data/llaviucu_pollen.csv") %>%
  select(depth,Charcoal.cc.) %>%
  left_join(pollen, by="depth") %>%
  select(depth, Charcoal.cc., upper_age, lower_age) %>%
  na.omit() 

#select human disturbance pollen taxa
agropastolarism_indicators <- c("Zea", "Hedyosmum", "Phaseolus", "Ipomoea", "Rumex", "Begoniaceae",
                                "Alnus", "Cyperaceae", "Cecropia", "Asteracea", "sporormiella")

# Create dataframe with agropastoralism taxa
agropastolarism <- select(pollen, contains(agropastolarism_indicators))
agropastolarism <- data.frame(sapply(agropastolarism, function(x) as.numeric((x))))
agropastolarism[is.na(agropastolarism)] <- 0 #Replace NA (if any) by 0
agropastolarism$charcoal <- charcoal$Charcoal.cc.

#drop agropastoralism indicators from pollen dataset
pollen <- pollen %>%  select(everything(), -contains(agropastolarism_indicators))

## extract agedepth variables
agedepth <- pollen[,names(pollen) %in% c("depth", "upper_age", "lower_age")]
pollen <- pollen[,!(names(pollen) %in% c("upper_age", "lower_age", "depth"))]

pollen <- data.frame(sapply(pollen, function(x) as.numeric((x))))
pollen[is.na(pollen)] <- 0 #Replace NA (if any) by 0

## assign dataframe to analyze
llaviucu_pollen <- pollen

##Calculate relative abundance (only for pollen)
total <- apply(llaviucu_pollen, 1, sum)
llaviucu_pollen <- llaviucu_pollen / total * 100

#Select spp (only for pollen)
abund <- apply(llaviucu_pollen, 2, max)
n.occur <- apply(llaviucu_pollen>0, 2, sum)
llaviucu_pollen <- llaviucu_pollen[, abund>1 & n.occur >2] #more than 1% of RA and present in >2 samples

#Save pollen data
#write.csv(llaviucu_pollen, "data/pollenRA.csv")

# Transform data to Hellinger form
core_hell <- decostand(llaviucu_pollen, method="hellinger")

# Run Principal Curves
pollen.prc <- prcurve(core_hell, method = "ca", trace = TRUE, vary = TRUE, penalty = 1.4)

## Extract position on the curve
cores_prc <- scores(pollen.prc, display = "curve")

# Combine dataframe with ages and depths
pollenPrC <- cbind(agedepth, cores_prc)

# Plot Pcurves with depth and ages
pollenPlot <- ggplot(pollenPrC, aes(x = upper_age, y = PrC)) +
  geom_line() + geom_point() +
  labs(y = "Pollen PrC", x = "Age (cal yr BP)", title = "") +
  ggtitle("Pollen PrC")
pollenPlot

# Save results
#write.csv(pollenPrC, "outputs/pollen-PrC.csv", row.names = FALSE)

## assign dataframe to analyze
llaviucu_pollen <- agropastolarism

# Transform data to Hellinger form
core_hell <- decostand(llaviucu_pollen, method="hellinger")

# Adjust for agropastoralism 
n.occur <- apply(core_hell>0, 1, sum)
core_hell2 <- core_hell[n.occur>0, ] #one single row has rowSums=0; remove it

# Run Principal Curves and extract scores for then combining the resulting dataframe with agedepth
pollen.prc <- prcurve(core_hell2, method = "ca", trace = TRUE, vary = TRUE, penalty = 1.4)
cores_prc <- scores(pollen.prc, display = "curve")

#AgropastPrC <- cbind(agedepth[-nrow(agedepth),], cores_prc)
AgropastPrC <- cbind(agedepth, cores_prc)

# Plot Pcurves with depth and ages
AgropastPlot <- ggplot(AgropastPrC, aes(x = upper_age, y = PrC)) +
  geom_line() + geom_point() +
  labs(y = "Agropastoralism PrC", x = "Age (cal yr BP)", title = "") +
  ggtitle("Agropastoralism PrC")
AgropastPlot

```

## Step 3: Interpolation of diatom dataset into pollen record
Diatom and pollen records are sampled from two different sediment cores with different temporal resolutions: 40-year resolution for pollen and Titanium, and 25-year resolution for diatoms. In order to facilitate comparison, we matched ages of pollen samples (coarser resolution) into diatom record (finer). The final dataset includes a common age column containing Principal curves for diatom, pollen, and agropastoralism, and Titanium time series. 

```{r interpolation, echo=TRUE, error=FALSE, message=FALSE, warning=FALSE, size="small"}
# read in XRF Llaviucu data (pollen record) and remove NAs
llaviucu_xrf <- read.csv("data/XRFTransformed3000yrs.csv") %>% rename(age=age_calBP) 
Ti_xrf <- na.omit(llaviucu_xrf[, c("age", "Ti")])

# read pollen data (drop Lycopodium counts)
pollen <- read.csv("data/llaviucu_pollen_counts.csv") %>%
  select(everything(), -contains("Lycopodium"))

#read in Charcoal data
charcoal <- read.csv("data/llaviucu_pollen.csv") %>%
  select(depth,Charcoal.cc.) %>%
  left_join(pollen, by="depth") %>%
  select(depth, Charcoal.cc., upper_age, lower_age) %>%
  na.omit() %>%
  mutate(zscores=(Charcoal.cc. - mean(Charcoal.cc.))/sd(Charcoal.cc.))


## read in PrC datasets: diatom, pollen and agropastoralism
diatomPrC <- readRDS("outputs/PrC-cores-diatoms.rds") %>%
  filter(Lake=="llaviucu") %>%
  mutate(elapsedTime = abs(upper_age - lower_age)) 

## Pollen PrC
pollenPrC <- read.csv("outputs/pollen-PrC.csv") %>%
  mutate(Age=upper_age) %>%
  mutate(elapsedTime = abs(upper_age - lower_age)) 
pollenPrC <- pollenPrC[order(pollenPrC$Age),] #order time

## Agropastoralism PrC
agropastPrC <- read.csv("outputs/agropastoralism-PrC.csv") %>%
  mutate(Age=upper_age) %>%
  mutate(elapsedTime = abs(upper_age - lower_age)) 
agropastPrC <- agropastPrC[order(agropastPrC$Age),] #order time

#Arrange diatom PrC in time order
diatomPrC <- diatomPrC[order(diatomPrC$Age),] #order time
diatomPrC <- na.omit(diatomPrC[, c("Age", "PrC", "elapsedTime")])
diatomPrC$Age <- round(diatomPrC$Age, 1)
mean(diff(diatomPrC$Age))


# linear interpolation diatom to pollen PrCs
diatom_to_pollen_prc_i <- as.data.frame(approx(diatomPrC$Age, diatomPrC$PrC, pollenPrC$Age)$y)
diatom_to_pollen_prc_i$age <- pollenPrC$Age
colnames(diatom_to_pollen_prc_i) <- c("PrC", "age")
diatom_to_pollen_prc_i$pollen_prc <- pollenPrC$PrC
#head(diatom_to_pollen_prc_i)
diff(diatom_to_pollen_prc_i$age)
diatom_to_pollen_prc_i$elapsedTime <- pollenPrC$elapsedTime
#plot(diatom_to_pollen_prc_i[,1], diatom_to_pollen_prc_i[,3])

# linear interpolation diatom to agropastoralism PrCs
diatom_to_agropast_prc_i <- as.data.frame(approx(diatomPrC$Age, diatomPrC$PrC, agropastPrC$Age)$y)
diatom_to_agropast_prc_i$age <- agropastPrC$Age
colnames(diatom_to_agropast_prc_i) <- c("PrC", "age")
diatom_to_agropast_prc_i$agropast_prc <- agropastPrC$PrC
#head(diatom_to_agropast_prc_i)
diff(diatom_to_agropast_prc_i$age)
diatom_to_agropast_prc_i$elapsedTime <- agropastPrC$elapsedTime
#plot(diatom_to_agropast_prc_i[,1], diatom_to_agropast_prc_i[,3])


# linear interpolation 2009 pollen ~ 2009 Ti--this is the same for agropast
Ti_pollen_prc_i <- as.data.frame(approx(Ti_xrf$age, Ti_xrf$Ti, pollenPrC$Age)$y)
Ti_pollen_prc_i$age <- pollenPrC$Age
colnames(Ti_pollen_prc_i) <- c("Ti", "age")
Ti_pollen_prc_i$pollen_prc <- pollenPrC$PrC
Ti_pollen_prc_i$elapsedTime <- pollenPrC$elapsedTime
#plot(Ti_pollen_prc_i[,1], Ti_pollen_prc_i[,3])


# Merge all datasets coarser dataset (pollen)
interpolatedData <- as.data.frame(cbind(diatom_to_pollen_prc_i$age, diatom_to_pollen_prc_i$PrC,
                                        diatom_to_pollen_prc_i$elapsedTime, #elapsed time in equal time resolution than Ti
                                        diatom_to_pollen_prc_i$pollen_prc, 
                                        diatom_to_agropast_prc_i$agropast_prc,
                                        Ti_pollen_prc_i$Ti))

colnames(interpolatedData) <- c("Age", "diatPrC", "elapsedTime_ti", "pollenPrC", "agropastPrC", "Ti")
median(diff(interpolatedData$Age))
head(interpolatedData)

# save data
#write.csv(interpolatedData, "outputs/principalcurves_ti_interp_coarser.csv")
```

## Visualization of Lake Llaviucu proxies
Here we generate an stacked plot of the interpolated time series
```{r summaryPlots, echo=TRUE, warning=FALSE}
# read data
df <- interpolatedData <- read.csv("outputs/principalcurves_ti_interp_coarser.csv")[,-1] 

# Define theme
theme_set(theme_bw())

# Plot stacked time series
data_plt <- df %>% 
  gather(key = proxy, value = var, -Age, -elapsedTime_ti)

plt <- ggplot(data_plt, aes(x=Age, y=var))+
  geom_line() +
  facet_grid(vars(proxy), scales = "free") +
  labs(y = "", x = "Age (cal years BP)", title = "")
plt

# plt_agropast <- ggplot(agroast_long, aes(x=upper_age, y=value))+
#   geom_line() +
#   facet_wrap(vars(taxa), scales = "free") +
#   theme(axis.title.y=element_text(size=12))+
#   labs(y = "", x = "Age (cal yr BP)", title = "")
# plt_agropast


# Save plot
# ggsave("outputs/stackedTS.png",
#        plot = plt_agropast,
#        width = 10,
#        height=8,
#        units="in",
#        dpi = 400)

```

# Step 4: Fit PrC-GAMs--**Covariates model**
In this step, we model diatom PrC trend as function of pollen PrC, agropastoralism PrC, and Titanium with Generalized Additive Models (GAMs) using the previous generated interpolated dataset. We then extract predicted values to test the temporal contribution of these covariates to explain drivers of diatom change over time.

```{r prc-gam, error=FALSE, message=FALSE, size="small"}
# Read in the interpolated dataset of diatoms, agropastoralism, and Titanium on pollen record
interpolatedData <- read.csv("outputs/principalcurves_ti_interp_coarser.csv")

# this chunk run the model weighting in the elapsed time of the pollen record
mod1 <- gam(diatPrC ~ s(Age, k=20, bs="ad") + s(Ti, k=10, bs="ad") + s(pollenPrC, k=15, bs="ad") + s(agropastPrC, k=15, bs="ad"),
            data = interpolatedData, method = "REML", 
            weights = elapsedTime_ti / mean(elapsedTime_ti),
            select = TRUE, family = gaussian(link="identity"),
            na.action = na.omit,
            knots = list(negAge=quantile(interpolatedData$Age, seq(0,1, length=10)))) #places notes at the deciles of sample ages

plot(mod1, page=1, scale = 0)
gam.check(mod1)
appraise(mod1)

## Predict GAM
# data to predict at. First check NAs and remove rows
row.has.na <- apply(interpolatedData, 1, function(x){any(is.na(x))})
interpolatedData <- interpolatedData[!row.has.na,]

## Predict
predGam <- cbind(interpolatedData, 
                 data.frame(predict.gam(mod1, interpolatedData, 
                 type = "terms" , se.fit = TRUE)))

#plot
var <- predGam$fit.s.agropastPrC.
se.var <- predGam$se.fit.s.agropastPrC.

predGamPlt <- ggplot(predGam, aes(x = Age, y = var)) +
  geom_line() +
  geom_ribbon(aes(ymin = var + (2 * se.var), ymax = var - (2 * se.var)),alpha=0.4) +
  geom_point()+
  #scale_x_reverse() +
  labs(y = "Agropastoralism", x = "cal years BP", title = "")+
  theme(legend.position = "none")+
  geom_hline(yintercept=0, linetype="dashed")+
  theme_bw()+ theme(legend.position = "none")+
  #ggtitle(expression("Agropastoralism" %->% "Diatoms")) +
  theme(strip.text.x = element_text(size = 12),
        axis.ticks.x = element_blank(),
        axis.text.y = element_text(size=12),
        axis.title.x = element_blank(),
        axis.text.x = element_text(size=12),
        axis.title.y=element_text(size=14))

var2 <- predGam$fit.s.pollenPrC.
se.var2 <- predGam$se.fit.s.pollenPrC.

predGamPlt2 <- ggplot(predGam, aes(x = Age, y = var2)) +
  geom_line() +
  geom_ribbon(aes(ymin = var2 + (2 * se.var2), ymax = var2 - (2 * se.var2)),alpha=0.4) +
  geom_point()+
  #scale_x_reverse() +
  labs(y = "Pollen", x = "cal years BP", title = "")+
  theme(legend.position = "none")+
  geom_hline(yintercept=0, linetype="dashed")+
  theme_bw()+ theme(legend.position = "none")+
  #ggtitle(expression("Pollen" %->% "Diatoms")) +
  theme(strip.text.x = element_text(size = 12),
        axis.ticks.x = element_blank(),
        axis.text.y=element_text(size=12),
        axis.title.x = element_blank(),
        axis.text.x = element_text(size=12),
        axis.title.y=element_text(size=14))

var3 <- predGam$fit.s.Ti.
se.var3 <- predGam$se.fit.s.Ti.

predGamPlt3 <- ggplot(predGam, aes(x = Age, y = var3)) +
  geom_line() +
  geom_ribbon(aes(ymin = var3 + (2 * se.var3), ymax = var3 - (2 * se.var3)),alpha=0.4) +
  geom_point()+
  #scale_x_reverse() +
  labs(y = "Titanium", x = "cal years BP", title = "")+
  theme(legend.position = "none")+
  geom_hline(yintercept=0, linetype="dashed")+
  theme_bw()+ theme(legend.position = "none")+
  #ggtitle(expression("Titanium" %->% "Diatoms")) +
  theme(strip.text.x = element_text(size = 12),
        axis.text.y=element_text(size=12),
        axis.text.x = element_text(size = 11),
        axis.title.x = element_text(size=12),
        axis.title.y=element_text(size=14))

# Make composite plot (Figure 3 of the manuscript)
library(cowplot)
predGam_all <- plot_grid(predGamPlt, predGamPlt2, predGamPlt3, align = "hv", 
                            nrow=3,ncol = 1, axis="tblr", labels = "AUTO")
predGam_all

# ggsave("outputs/diatomPrC_GAM_covariates.png",
#        plot = predGam_all ,
#        width = 8,
#        height=10,
#        units="in",
#        dpi = 150)

```

## Step 5: Hierarchical Generalized Additive Models (HGAM) -- **Synchronous model**
In this step we model, first, diatom and pollen assemblages using HGAM and then calculate average rate of change at assemblage-level to assess the timing in the start of rapid change in the time-series, and test comparable rates of change between aquatic and terrestrial ecosystems.

By obtaining derivative-based values of the diatom, terrestrial pollen, and agropastoralism time series we obtain regular time intervals at 24-year resolution (median between diatom and pollen records) and hence comparable--same time steps-- for fitting the synchronous and asynchronous models. 

We fit model I HGAM (Pedersen et al., 2019): group-level smoothness between groups (spp) without a global smooth while allowing different willingness to not assume any shape among species for each assemblage.

### Diatom assemblages
This chunk is reproducible for pollen and agropastoralism assemblages

```{r HGAMdiatoms, echo=TRUE, error=FALSE, message=FALSE, size="small"}
#read dataframe with diatom absolute counts
mergedCores <- read.csv("data/mergedCores_counts4.csv")[-1] 

#import dataframe wiht old and new names to group
changes <- read.csv("data/old_new_nms_cores_counts.csv", stringsAsFactors = FALSE)
#new1: ecological groups
#new2: harmonized taxonomic names

#filter cores
select <- c("llaviucu")

# this is to select the most common species for a single lake   
core <- mergedCores %>% 
  filter(str_detect(lake, select))

agedepth <- core[, names(core) %in% c("depth", "upper_age", "lower_age", "lake")]
diat <- core[, !names(core) %in% c("depth", "upper_age", "lower_age", "lake")]
diat[is.na(diat)] <- 0

#Select most common species 
criteria <- 0.5 #% of the total samples

n.occur <- apply(diat>0, 2, sum)
diat_red <- diat[, n.occur > (dim(diat)[1])*criteria] #
diat <- cbind(agedepth, diat_red)

##This creates the dataset containing the most common species for a single lake core
#Gather
spp_thin <- diat %>% 
  gather(key = taxa, value = count, -depth, -upper_age, -lower_age,  -lake) #don't gather depths, ages and lake variables

diat_data <- spp_thin %>%
  mutate(taxa = plyr::mapvalues(taxa, from = changes$old, to = changes$new_2)) %>%
  group_by(depth) %>%
  mutate(total_sample = sum(count)) %>% 
  filter(!total_sample == "0") %>% #this is to remove empty samples
  filter(!upper_age == 0) %>% #this is to remove ages == 0 (triumfo and fondodocha record)
  mutate(log_total_counts = log10(total_sample+1)) %>%
  mutate(relative_abundance_percent = count / sum(count) * 100) %>%
  mutate(negAge = -upper_age) %>%
  mutate(AgeCE = upper_age*(-1)+1950) %>%
  #filter(AgeCE >= "0") %>%
  mutate(elapsedTime = abs(upper_age - lower_age)) %>%
  ungroup() %>%
  group_by(lake) %>%
  mutate(spp = factor(taxa)) 

levels(diat_data$spp) #check how many diatom species

set.seed(10) #set a seed so this is repeatable

#model I HGAM: different smoothness for each taxa without a global smooth
diatom_gam_I<- gam(count ~ s(negAge, by=spp, k=20, bs="fs") +
                     s(spp, bs="re") + offset(log_total_counts),
                   weights = elapsedTime / mean(elapsedTime),
                   data=diat_data, family = nb,
                   knots = list(negAge=quantile(diat_data$negAge, seq(0,1, length=10))), #places notes at the deciles of sample ages
                   method = "REML")
gam.check(diatom_gam_I)
```

#### Create synthetic data to predict over a range of ages and plot group-level trends with different smootheness (HGAM)

```{r HGAMpredict, error=FALSE, message=FALSE, size="small"}
diat_plot_data <- with(diat_data, as_tibble(expand.grid(negAge = seq(min(diat_data$negAge), max(diat_data$negAge)),
                                                        spp = factor(levels(diat_data$spp)),
                                                        log_total_counts = mean(log_total_counts))))

diat_modI_fit <- predict(diatom_gam_I,
                         newdata = diat_plot_data,
                         se.fit = TRUE)

#non-shared trends
diat_plot_data$modI_fit <- as.numeric(diat_modI_fit$fit)

# For only one model
diat_plot_data <- gather(diat_plot_data, key=model, value=fit, modI_fit)
diat_plot_data <- mutate(diat_plot_data, se= c(as.numeric(diat_modI_fit$se.fit)),
                         upper = exp(fit + (2 * se)),
                         lower = exp(fit - (2 * se)),
                         fit   = exp(fit))

#non-shared trends
theme_set(theme_bw())
theme_update(panel.grid = element_blank())

diat_plot <- ggplot(diat_plot_data) +
  facet_wrap(~spp, nrow = 4,scales = "free_y")+
  geom_ribbon(aes(x=negAge,
                  ymin = lower,
                  ymax = upper,
                  fill = model),
              alpha=0.2)+
  scale_x_reverse()+
  geom_point(data= diat_data, aes(x = negAge, y = count), size=0.06) +
  geom_line(aes(x = negAge, y = fit, color = model))+
  labs(y = "Absolute counts", x = "Age (cal yr BP)") +
  ggtitle("HGAM diatoms") +
  scale_fill_brewer(name = "", palette = "Dark2") +
  scale_colour_brewer(name = "",
                      palette = "Dark2")+
  theme(legend.position = "top",
        strip.text = element_text(size=10))

diat_plot
```
## Step 6: Calculation of derivatives and posterior distribution simulation
```{r derivatives, error=FALSE, message=FALSE, size="small"}
set.seed(10) #set a seed so this is repeatable
n_sims = 250
n_length = 140
years <- seq(min(diat_plot_data$negAge),
             max(diat_plot_data$negAge),
             length.out = n_length)

model <- diatom_gam_I
pred <- diat_modI_fit

# Generate multivariate normal simulations of model coefficients
random_coefs <- t(rmvn(n_sims, mu = coef(model),V = vcov(model)))

confint_sims <- crossing(spp=unique(diat_plot_data$spp),
                         negAge = seq(min(diat_plot_data$negAge),
                                      max(diat_plot_data$negAge),
                                      length.out = n_length),
                         log_total_counts=0)

map_pred_sims <- predict(model,
                         confint_sims,
                         type = "lpmatrix") %*% random_coefs %>%
  as_data_frame() %>%
  bind_cols(confint_sims)%>%
  gather(key = simulation, value = pred, -negAge, -log_total_counts,-spp)


#specifying the step size for numerical derivative calculations
delta = 0.01

#calculating the predicted value for the current year plus delta
step_ahead_fits = confint_sims %>%
  mutate(negAge = negAge+delta)%>%
  predict(model, 
          ., type = "lpmatrix") %*% random_coefs 


#calculating the predicted value for the current year minus delta
step_behind_fits = confint_sims %>%
  mutate(negAge = negAge-delta)%>%
  predict(model,
          ., type = "lpmatrix") %*% random_coefs 


#Function for calculating first derivatives of time series given a before,
#after, and delta step size
calc_1st_deriv = function(fit_before, fit_after,delta) {
  (fit_after-fit_before)/(2*delta)
}


#using the predicted values for year plus and minus delta to calculate
#derivatives for each species for each simulation

derivs <- calc_1st_deriv(step_behind_fits,step_ahead_fits,delta = delta)%>%
  as_data_frame()%>%
  bind_cols(confint_sims)%>%
  gather(key = simulation,value = deriv, -spp,-negAge, -log_total_counts)

#Creating summaries of derivatives for each simulation for each year
deriv_summaries <- derivs %>%
  group_by(negAge,simulation)%>%
  summarize(deriv_mean = mean(deriv),
            deriv_sd = sd(deriv))%>%
  group_by(negAge)%>% #turning derivative summaries into 95% confidence intervals
  select(-simulation)%>%
  summarize_all(.funs = list(lower = ~quantile(.,probs = 0.025),
                             upper = ~quantile(.,probs = 0.975),
                             med   = ~quantile(.,probs = 0.5)))

diatoms_deriv <- deriv_summaries

```

#### Plotting the HGAM-derivative diatom time series

```{r derivatives plot, error=FALSE, message=FALSE, size="small"}
#Plotting mean rate of change plus the 95% CI
mean_plot <- diatoms_deriv %>%
  ggplot(aes(x = negAge, 
             y = deriv_mean_med, 
             ymin = deriv_mean_lower,
             ymax = deriv_mean_upper))+
  geom_ribbon(fill="grey")+
  geom_line()+
  geom_hline(yintercept = 0, linetype=2) +
  scale_y_continuous("")+
  xlab("Cal years BP") +
  theme_bw()
mean_plot 

#Plotting standard deviation of rate of change plus the 95% CI
sd_plot <- deriv_summaries %>%
  ggplot(aes(x = negAge, 
             y = deriv_sd_med, 
             ymin=deriv_sd_lower,
             ymax=deriv_sd_upper))+
  geom_ribbon(fill="grey")+
  geom_line()+
  geom_hline(yintercept = 0, linetype=2) +
  scale_y_continuous("")+
  xlab("Cal years BP") +
  theme_bw()
sd_plot
```

## Step 7: Generalized Least Squares (GLS) -- **Asynchronous model**

The **asynchronous model** consists in two steps:

+  Generating lagged time series: diatom samples (response) are paired with antecedent samples of predictors (pollen and agropastoralism)
+  Modeling time-delayed effects of pollen and agropastoralism rates of change (separately) on diatom rate of change

We use functions developed by **Blas Benito** in his co-authored paper led by **Graciela Gil-Romera** and colleagues [Long-term fire resilience of the Ericaceous Belt, Bale Mountains, Ethiopia](https://github.com/BlasBenito/BaleFire)

Given the temporal mean age resolution of the diatom record (ca 25 years), we generated 40 lags (2500 years matching the two record lengths). Every diatom sample has been paired with the antecedent pollen and agropastoralism samples (separately). This process is repeated for every sample and every lag.

The question to be answered by applying this model is: Are time-delayed effects of pollen (and agropastoralism) rates of change on current rate of change of diatoms? This is written in the following Equation 1:

**Equation 1**: $$DiatomRateofChange_{t} = \alpha + \beta Predictors_{t+lag} + \epsilon$$

Where:
+  $t$ is the age of the diatom sample
+  $lag$ is the time-span in between the response samples and the antecedent samples of the predictors (pollen and agropastoralism.

The function **backwardLags** aims to assess the effect of “past” pollen or agropastoralism values on diatoms.
The function *modelLagData* fits one Generalized Least Squares (GLS) model per lag, and computes pseudo R-squared values and standardized coefficient estimates.



```{r GLS, error=FALSE, message=FALSE, size="small"}
# Read in smoothed derivative proxy (diatoms, native pollen, and agropastoralism) time series
derivAll <- read.csv("outputs/pollen-agropast-diatom-deriv_24timeSteps.csv")

# Read in functions
source("scripts/functions_custom.R")

# Subset datasets: diatoms
diat <- derivAll %>% filter(proxy=="diatoms") %>% select(negAge,deriv_mean_med) %>%
  mutate(age=negAge) %>% data.frame()
diat$age <- round(diat$age, 1)
diat$negAge <- NULL
colnames(diat) <- c("diatom_deriv", "age")
round(diff(diat$age, 1))[1]

# Subset datasets: Native pollen
pollen <- derivAll %>% filter(proxy=="pollen") %>% select(negAge, deriv_mean_med) %>%
  mutate(age=negAge) %>% data.frame()
pollen$age <- round(pollen$age, 1)
pollen$negAge <- NULL
colnames(pollen) <- c("pollen_deriv", "age")
round(diff(pollen$age, 1))[1] #check temporal resolution

# Subset datasets: Agropastoralism
agropastolarism <- derivAll %>% filter(proxy=="agropastoralism") %>% select(negAge, deriv_mean_med) %>%
  mutate(age=negAge) %>% data.frame()
agropastolarism$age <- round(agropastolarism$age, 1)
agropastolarism$negAge <- NULL
colnames(agropastolarism) <- c("pollen_deriv", "age") #note the change of the column name
round(diff(agropastolarism$age, 1))[1]

### Backward lags
lags<-1:40

#backward dataset 
#to assess the effect of “past” pollen or agropastoralism values on diatoms
lag.data.backward <- backwardLags(
  lags=lags, 
  reference.data=diat, 
  data.to.lag=pollen
)

# Fitting a GLS model per lag on backward datasets
backward.results <- modelLagData(
  model.formula="diatom_deriv ~ pollen_deriv", 
  lagged.data=lag.data.backward
)

backward.results$value <- round(backward.results$value, 2)

# Fitting a null model
backward.results.random <- modelRandomLagData(
  lagged.data=lag.data.backward, 
  model.formula="diatom_deriv ~ pollen_deriv", 
  iterations=1000
)

backward.results.random$value <- round(backward.results.random$value, 2)


# Plot model results
## Set axes limits
max.lag = max(c(backward.results$lag))
max.coefficient = round(max(c(backward.results[backward.results$variable=="Coefficient", "value"], backward.results.random[backward.results.random$variable=="Coefficient", "upper"])) + 0.1, 1)
min.coefficient = round(min(c(backward.results[backward.results$variable=="Coefficient", "value"], backward.results.random[backward.results.random$variable=="Coefficient", "lower"])) - 0.1, 1)
max.R2 = round(max(c(backward.results[backward.results$variable=="R2", "value"])), 1)

library(viridis)
viridis.colors <- viridis(10, option="D")

backward.plot.coefficient <- ggplot(data=subset(backward.results, variable=="Coefficient"), aes(x=lag, y=value)) +
  geom_ribbon(data=subset(backward.results.random, variable=="Coefficient"), aes(ymin=lower, ymax=upper), alpha=0.3, fill="light grey") +
  geom_line(data=subset(backward.results.random, variable=="Coefficient"), aes(x=lag, y=value), alpha=0.6, color="light grey", size=1) +
  geom_hline(yintercept=0, color="black", linetype=2) +
  geom_ribbon(aes(ymin=lower,ymax=upper), alpha=0.3, fill=viridis.colors[2]) +
  geom_line(size=1.5, color=viridis.colors[1]) +
  ggtitle(expression("Pollen" %->% "Diatoms")) +
  theme(legend.position="none") +
  xlab("") +
  ylab("Standardized coefficient") +
  scale_y_continuous(breaks=seq(min.coefficient, max.coefficient, by=0.8)) +
  #scale_x_reverse()+
  theme(axis.text.y = element_text(size=16),
        axis.text.x = element_text(size=16),
        axis.title.y = element_text(size=16),
        plot.title = element_text(size=16),
        plot.margin = unit(c(0, 0, 0, 0), "cm"),
        axis.title.x=element_blank()) +
        #axis.text.x=element_blank(),
        #axis.line.x=element_blank(),
        #axis.ticks.x = element_blank()) +
  theme_classic()
  #coord_cartesian(ylim = c(min.coefficient, max.coefficient + 0.5))

#backward.plot.coefficient

backward.plot.R2 <- ggplot(data=subset(backward.results, variable=="R2"), aes(x=lag, y=value, group=variable)) +
  geom_ribbon(data=subset(backward.results.random, variable=="R2"), aes(ymin=lower, ymax=upper), alpha=0.3, fill="light grey") +
  geom_line(data=subset(backward.results.random, variable=="R2"), aes(x=lag, y=value), alpha=0.6, color="light grey", size=1.5) +
  geom_line(size=1.5, color=viridis.colors[2]) +
  theme(legend.position="none") +
  xlab("Years (before Diatom samples)") +
  ylab("Pseudo R squared") +
  scale_y_continuous(breaks=seq(0, max.R2, by=0.1)) +
  #scale_x_reverse()+
  theme(axis.text.y = element_text(size=16),
        axis.text.x = element_text(size=16),
        plot.title = element_text(size=16),
        axis.title.y = element_text(size=16),
        plot.margin = unit(c(0.2, 0.5, 0, 0), "cm")) +
  theme_classic() +
  coord_cartesian(ylim = c(0, max.R2 + 0.05))

#backward.plot.R2

#Combine the two plots
library(cowplot)
plot_composite <- plot_grid(backward.plot.coefficient, 
                            backward.plot.R2, ncol = 1, 
                            rel_heights = c(1, 1), align="v") + 
                            theme(plot.margin = unit(c(0.5, -1, 0.5, 0.5), "cm"))
plot_composite

# save plot
ggsave("outputs/pollen_diatoms_asycnchronousModel.png",
       plot = plot_composite,
       width=8,
       height=6,
       units="in",
       dpi = 300)




```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
