---
title: "Paper Diatom-Pollen-Coupling"
author: "XBenito"
date: "15/4/2021"
output: word_document
---

```{r setup, echo=FALSE, message=FALSE, error=FALSE, warning=FALSE, results="hide"}
knitr::opts_chunk$set(echo = TRUE)
```

## Workflow of the main numerical analyses. This document shows the most relevant code. See R files within the folder "scripts" for a more specific code.

# Study area and datasets description
**Lake Llaviucu**: morrained-dammed tropical Lake located in Ecuador (Cajas National Park), at 3115 m of elevation.
Two sediment cores: 
  - **Diatom**: 3-m core covering the last ca 3,000 years, and sampled at 1-cm intervals for diatom analysis [Benito et al. 2021 L&O] (https://aslopubs.onlinelibrary.wiley.com/doi/full/10.1002/lno.11747) 
  - **Pollen**: 11.9-m core covering the last 14,000 years, and sampled at 1-cm intervals for pollen, charcoal and Sporiormella analyses [Nascimiento et al. 2020 QSR] (https://www.sciencedirect.com/science/article/abs/pii/S0277379120304339?via%3Dihub)


# Step 1: Estimate Principal Curves of the diatom record
```{r}
library(analogue)
library(tidyverse)

#read diatom core datasets
mergedCores <- read.csv("data/mergedCores_counts4.csv") #this is a dataframe with absolute counts 

agedepth <- mergedCores[, names(mergedCores) %in% c("depth", "upper_age", "lower_age", "lake")]
diat <- mergedCores[, !names(mergedCores) %in% c("depth", "upper_age", "lower_age", "lake")]
diat[is.na(diat)] <- 0

diatoms_save <- cbind(agedepth, diat)

changes <- read.csv("data/old_new_nms_cores_counts.csv", stringsAsFactors = FALSE)
#new1: ecological groups
#new2: harmonized taxonomic names

#this is to transform to tidy format, calculate % and subset more common species
new <- diatoms_save %>% 
  gather(key = taxa, value = count, -depth, -upper_age, -lower_age, -lake) %>%
  mutate(taxa = plyr::mapvalues(taxa, from = changes$old, to = changes$new_2)) %>%
  group_by(depth, taxa, lake, upper_age, lower_age) %>%
  summarise(count = sum(count)) %>%
  filter(!count == "0" ) %>% #this is to remove empty samples (rows)
  filter(!upper_age==0.0) %>% #this is to drop extraneous ages
  ungroup() %>%
  group_by(depth, lake) %>%
  mutate(relative_abundance_percent = count / sum(count) * 100) %>%
  ungroup()

# filter more abundant taxa; format need to be on long-wide format-->no spreaded 
core_common_taxa <- new %>%
  group_by(taxa) %>%
  summarise(max_rel_abund = max(relative_abundance_percent)) %>%
  filter(max_rel_abund >= 5) %>%
  arrange(max_rel_abund) %>%
  pull(taxa)

# select from initial table
core_counts_common <- new %>%
  filter(taxa %in% core_common_taxa) %>%
  mutate(taxa = factor(taxa, levels = core_common_taxa)) %>%
  arrange(taxa)

#make it wide
core_counts_wide <- core_counts_common %>%
  select(depth, lake, upper_age, lower_age, taxa, relative_abundance_percent) %>%
  spread(key = taxa, value = relative_abundance_percent) 

## split cores by lakes and reassemble
coresList <- split(core_counts_wide, core_counts_wide$lake)

# Extract LLaviucu diatom dataset
llaviucu_diatoms <- as.data.frame(coresList[["llaviucu"]])
llaviucu_diatoms[is.na(llaviucu_diatoms)] <- 0

## extract agedepth variables
agedepth <- llaviucu_diatoms[,names(llaviucu_diatoms) %in% c("depth", "lake", "upper_age", "lower_age")]
llaviucu_diatoms <- llaviucu_diatoms[,!names(llaviucu_diatoms) %in% c("depth", "lake", "upper_age", "lower_age")]

#Select most abundant and common species
abund <- apply(llaviucu_diatoms, 2, max)
n.occur <- apply(llaviucu_diatoms>0, 2, sum)
llaviucu_diatoms <- llaviucu_diatoms[, abund>1 & n.occur >2] #more than 1% of RA and present in >2 samples

# Combine dataframe with ages and depths
llaviucu_diatoms <- cbind(agedepth, llaviucu_diatoms)

# Save Diatom data RA
write.csv(llaviucu_diatoms, "data/diatomsRA.csv")

# Function to fit principal curves
fitPcurve <- function(i, cores, axis, method, ...) {
  axis <- axis[i]
  core <- cores[[i]]
  method <- method[i]
  core <- core[ , -which(names(core) %in% c("depth","upper_age", "lower_age", "lake"))] # drop year & depths vars
  core[is.na(core)] <- 0
  core <- core[, colSums(core) > 0] #select only present species
  core <- decostand(core, method="hellinger") #Hellinger transform relative abundance data
  prc <- prcurve(core, vary = TRUE, trace = TRUE, method = method,
                 axis = axis)
  prc
}

## apply fitCurve  function to each core
axis <- c(1, 1, 1, 1, 1, 1, 1, 1) 
method <- c(rep("ca", 8)) 

corePRC <- lapply(seq_along(coresList), fitPcurve,
                  cores = coresList, axis = axis, method = method)
names(corePRC) <- names(coresList)

# extract data from list
scrs <- lapply(corePRC, scores)
scrs <- lapply(scrs, `[`, , 1)
len <- lapply(scrs, length)
nams <- names(scrs)
depths <- lapply(coresList, `[`, , "depth")
age <- lapply(coresList, `[`, ,  "upper_age")
upper_age <- lapply(coresList, `[`, ,  "upper_age")
lower_age <- lapply(coresList, `[`, ,  "lower_age")
df <- data.frame(Lake = rep(nams, times = len),
                 PrC  = unlist(scrs),
                 Depth = unlist(depths),
                 Age = unlist(age),
                 negAge = - unlist(age),
                 upper_age = unlist(upper_age),
                 lower_age = unlist(lower_age))
rownames(df) <- NULL

#Plot Llaviucu diatom PrC
cores.plot <- ggplot(df[df$Lake=="llaviucu",], aes(x = Age, y = PrC)) +
  geom_point() + geom_line() +
  #coord_flip() +
  facet_wrap(Lake~., scales = "free")
cores.plot

## write this out for use in the GAM modelling script
saveRDS(df, file = "outputs/PrC-cores-diatoms.rds") #this is including all spp

```

# Step 2: Estimate Principal Curves of pollen record
```{r principal curves, size="small", results="hide", message=FALSE, error=FALSE, warning=FALSE, echo="FALSE"}

# read Pollen data (drop Lycopodium counts)
pollen <- read.csv("data/llaviucu_pollen_counts.csv") %>%
  select(everything(), -contains("Lycopodium"))

#read in Charcoal data
charcoal <- read.csv("data/llaviucu_pollen.csv") %>%
  select(depth,Charcoal.cc.) %>%
  left_join(pollen, by="depth") %>%
  select(depth, Charcoal.cc., upper_age, lower_age) %>%
  na.omit() 

#select human disturbance pollen taxa
agropastolarism_indicators <- c("Zea", "Hedyosmum", "Phaseolus", "Ipomoea", "Rumex", "Begoniaceae",
                                "Alnus", "Cyperaceae", "Cecropia", "Asteracea", "sporormiella")

# Create dataframe with agropastoralism taxa
agropastolarism <- select(pollen, contains(agropastolarism_indicators))
agropastolarism <- data.frame(sapply(agropastolarism, function(x) as.numeric((x))))
agropastolarism[is.na(agropastolarism)] <- 0 #Replace NA (if any) by 0
agropastolarism$charcoal <- charcoal$Charcoal.cc.

#drop agropastoralism indicators from pollen dataset
pollen <- pollen %>%  select(everything(), -contains(agropastolarism_indicators))

## extract agedepth variables
agedepth <- pollen[,names(pollen) %in% c("depth", "upper_age", "lower_age")]
pollen <- pollen[,!(names(pollen) %in% c("upper_age", "lower_age", "depth"))]

pollen <- data.frame(sapply(pollen, function(x) as.numeric((x))))
pollen[is.na(pollen)] <- 0 #Replace NA (if any) by 0

## assign dataframe to analyze
llaviucu_pollen <- pollen

##Calculate relative abundance (only for pollen)
total <- apply(llaviucu_pollen, 1, sum)
llaviucu_pollen <- llaviucu_pollen / total * 100

#Select spp (only for pollen)
abund <- apply(llaviucu_pollen, 2, max)
n.occur <- apply(llaviucu_pollen>0, 2, sum)
llaviucu_pollen <- llaviucu_pollen[, abund>1 & n.occur >2] #more than 1% of RA and present in >2 samples

# Combine dataframe with ages and depths
llaviucu_pollen <- cbind(agedepth, llaviucu_pollen)

#Save pollen data
write.csv(llaviucu_pollen, "data/pollenRA.csv")

# Transform data to Hellinger form
core_hell <- decostand(llaviucu_pollen, method="hellinger")

# Run Principal Curves
pollen.prc <- prcurve(core_hell, method = "ca", trace = TRUE, vary = TRUE, penalty = 1.4)

## Extract position on the curve
cores_prc <- scores(pollen.prc, display = "curve")

# Combine dataframe with ages and depths
pollenPrC <- cbind(agedepth, cores_prc)

# Plot Pcurves with depth and ages
pollenPlot <- ggplot(pollenPrC, aes(x = upper_age, y = PrC)) +
  geom_line() + geom_point() +
  labs(y = "Pollen PrC", x = "Age (cal yr BP)", title = "")

pollenPlot

# Save results
write.csv(pollenPrC, "outputs/pollen-PrC.csv", row.names = FALSE)

## assign dataframe to analyze
llaviucu_pollen <- agropastolarism

# Transform data to Hellinger form
core_hell <- decostand(llaviucu_pollen, method="hellinger")

# Adjust for agropastoralism 
n.occur <- apply(core_hell>0, 1, sum)
core_hell2 <- core_hell[n.occur>0, ] #one single row has rowSums=0; remove it

# Run Principal Curves and extract scores for then combining the resulting dataframe with agedepth
pollen.prc <- prcurve(core_hell2, method = "ca", trace = TRUE, vary = TRUE, penalty = 1.4)
cores_prc <- scores(pollen.prc, display = "curve")
#AgropastPrC <- cbind(agedepth[-nrow(agedepth),], cores_prc)
AgropastPrC <- cbind(agedepth, cores_prc)

## save results
write.csv(AgropastPrC, "outputs/agropastoralism-PrC.csv", row.names = FALSE)

# Plot Pcurves with depth and ages
AgropastPlot <- ggplot(AgropastPrC, aes(x = upper_age, y = PrC)) +
  geom_line() + geom_point() +
  labs(y = "Agropastoralism PrC", x = "Age (cal yr BP)", title = "")
AgropastPlot

```


# Step 3: Interpolation of diatom dataset into pollen record
Diatom and pollen records are sampled from two different sediment cores. In order to facilitate comparison we matched ages of diatom samples into pollen and Titanium (as indicator of terrigenous input) record. The final dataset includes a common age column containing Principal curves for diatom, pollen, and agropastoralism, and Ti time series. 

```{r interpolation, message=FALSE, error=FALSE, results="hide", size="small"}
library(tidyverse)

# read in XRF Llaviucu data (pollen record)
llaviucu_xrf <- read.csv("data/XRFTransformed3000yrs.csv") %>% rename(age=age_calBP) 
Ti_xrf <- na.omit(llaviucu_xrf[, c("age", "Ti")])

# read pollen data (drop Lycopodium counts)
pollen <- read.csv("data/llaviucu_pollen_counts.csv") %>%
  select(everything(), -contains("Lycopodium"))

#read in Charcoal data
charcoal <- read.csv("data/llaviucu_pollen.csv") %>%
  select(depth,Charcoal.cc.) %>%
  left_join(pollen, by="depth") %>%
  select(depth, Charcoal.cc., upper_age, lower_age) %>%
  na.omit() %>%
  mutate(zscores=(Charcoal.cc. - mean(Charcoal.cc.))/sd(Charcoal.cc.))


## read in PrC datasets: diatom, pollen and agropastoralism
diatomPrC <- readRDS("outputs/PrC-cores-diatoms.rds") %>%
  filter(Lake=="llaviucu") %>%
  mutate(elapsedTime = abs(upper_age - lower_age)) 

## Pollen PrC
pollenPrC <- read.csv("outputs/pollen-PrC.csv") %>%
  mutate(Age=upper_age) %>%
  mutate(elapsedTime = abs(upper_age - lower_age)) 
pollenPrC <- pollenPrC[order(pollenPrC$Age),] #order time

## Agropastoralism PrC
agropastPrC <- read.csv("outputs/agropastoralism-PrC.csv") %>%
  mutate(Age=upper_age) %>%
  mutate(elapsedTime = abs(upper_age - lower_age)) 
agropastPrC <- agropastPrC[order(agropastPrC$Age),] #order time

#Arrange diatom PrC in time order
diatomPrC <- diatomPrC[order(diatomPrC$Age),] #order time
diatomPrC <- na.omit(diatomPrC[, c("Age", "PrC", "elapsedTime")])
diatomPrC$Age <- round(diatomPrC$Age, 1)
mean(diff(diatomPrC$Age))

# linear interpolation diatom ~ Ti (pollen record)
Ti_diat_prc_i <- as.data.frame(approx(Ti_xrf$age, Ti_xrf$Ti, diatomPrC$Age)$y)
Ti_diat_prc_i$age <- diatomPrC$Age
colnames(Ti_diat_prc_i) <- c("Ti", "age")
Ti_diat_prc_i$diat_prc <- diatomPrC$PrC #add PrC data
diff(Ti_diat_prc_i$age) #check temporal resolution
Ti_diat_prc_i$elapsedTime <- diatomPrC$elapsedTime #add sample age

# linear interpolation diatom ~ charcoal (pollen record)
charc_diat_prc_i <- as.data.frame(approx(charcoal$upper_age, charcoal$Charcoal.cc., diatomPrC$Age)$y)
charc_diat_prc_i$age <- diatomPrC$Age
colnames(charc_diat_prc_i) <- c("charcoal", "age")
charc_diat_prc_i$diat_prc <- diatomPrC$PrC #add PrC data
diff(charc_diat_prc_i$age) #check temporal resolution


# linear interpolation pollen ~ Ti
Ti_pollen_prc_i <- as.data.frame(approx(Ti_xrf$age, Ti_xrf$Ti, pollenPrC$Age)$y)
Ti_pollen_prc_i$age <- pollenPrC$Age
colnames(Ti_pollen_prc_i) <- c("Ti", "age")
Ti_pollen_prc_i$pollen_prc <- pollenPrC$PrC #add PrC data
Ti_pollen_prc_i$elapsedTime <- pollenPrC$elapsedTime #check temporal resolution


# linear interpolation agropastoralism ~ Ti
Ti_agropast_prc_i <- as.data.frame(approx(Ti_xrf$age, Ti_xrf$Ti, agropastPrC$Age)$y)
Ti_agropast_prc_i$age <- agropastPrC$Age
colnames(Ti_agropast_prc_i) <- c("Ti", "age")
Ti_agropast_prc_i$agrospast_prc <- agropastPrC$PrC

# Merge all datasets
interpolatedData <- as.data.frame(cbind(Ti_diat_prc_i$age, Ti_diat_prc_i$Ti,
                          charc_diat_prc_i$charcoal,
                          Ti_pollen_prc_i$elapsedTime, #elapsed time in equal time resolution than Ti
                          Ti_diat_prc_i$elapsedTime, #elapsed time in equal time resolution than diatoms
                          Ti_diat_prc_i$diat_prc,
                          Ti_pollen_prc_i$pollen_prc, Ti_agropast_prc_i$agrospast_prc))

colnames(interpolatedData) <- c("Age", "Ti", "charcoal", "elapsedTime_ti", "elapsedTime_diat", "diatPrC", "pollenPrC", "agropastPrC")

# save data
write.csv(interpolatedData, "outputs/principalcurves_ti_interp.csv")
```

# Visualization of Lake Llaviucu proxies
Here we generate an stacked plot of the time series
```{r summaryPlots}
# read data
df <- interpolatedData <- read.csv("outputs/principalcurves_ti_interp.csv")[,-1] %>%
  mutate(sample_age_mean=(elapsedTime_ti+elapsedTime_diat)/2) #average sample ages of pollen and diatom records
head(df)

# Define theme
theme_set(theme_bw())

# Plot stacked time series
data_plt <- df %>% 
  gather(key = proxy, value = var, -Age, -elapsedTime_ti, -elapsedTime_diat)

plt <- ggplot(data_plt, aes(x=Age, y=var))+
  geom_line() +
  facet_grid(vars(proxy), scales = "free") +
  labs(y = "", x = "Age (cal yr BP)", title = "")
plt

# plt_agropast <- ggplot(agroast_long, aes(x=upper_age, y=value))+
#   geom_line() +
#   facet_wrap(vars(taxa), scales = "free") +
#   theme(axis.title.y=element_text(size=12))+
#   labs(y = "", x = "Age (cal yr BP)", title = "")
# plt_agropast


# Save plot
ggsave("outputs/stackedTS.png",
       plot = plt_agropast,
       width = 10,
       height=8,
       units="in",
       dpi = 400)

# Generate a series of phase-space for drivers (pollen, agropastoralism, Ti, charcoal) vs response variables (diatoms) coloured by year

ggplot(df, aes(x=pollenPrC, y=diatPrC, colour=Age)) +
  geom_point(size=2) +
  scale_color_viridis_c()

ggplot(df, aes(x=agropastPrC, y=diatPrC, colour=Age)) +
  geom_point(size=2) +
   scale_color_viridis_c()

ggplot(df, aes(x=Ti, y=diatPrC, colour=Age)) +
  geom_line(size=2) +
   scale_color_viridis_c()

ggplot(df, aes(x=charcoal, y=diatPrC, colour=Age)) +
  geom_point(size=2) +
   scale_color_viridis_c()

```


# Step 4: Fit PrC-GAMs 
In this step we model diatom PrC as function of pollen, agropastoralism and Ti with Generalized Additive Models (GAMs). We then extract predicted values to test the temporal contribution of these covariates to explain drivers of diatom assemblage change.

```{r prc-gam, message=FALSE, error=FALSE, results="hide", size="small"}
library(mgcv)
library(tidyverse)
library(ggplot2)
library(gratia)

# Read in the interpolated dataset of Ti, pollen and agropastoralism PrCs 
interpolatedData <- read.csv("outputs/principalcurves_ti_interp.csv")[,-1] %>%
  mutate(sample_age_mean=(elapsedTime_ti+elapsedTime_diat)/2) #average sample ages of pollen and diatom records

# this chunk run the model weighting in the elapsed time of the pollen record
mod1 <- gam(diatPrC ~ s(Age, k=40, bs="ad") + s(Ti) + s(pollenPrC) + s(agropastPrC),
            data = interpolatedData, method = "REML", 
            weights = elapsedTime_diat / mean(elapsedTime_diat),
            select = TRUE, family = gaussian(link="identity"),
            na.action = na.omit,
            knots = list(negAge=quantile(interpolatedData$Age, seq(0,1, length=10)))) #places notes at the deciles of sample ages

plot(mod1, page=1, scale = 0)
gam.check(mod1)
appraise(mod1)
summary(mod1)


#Check temporal autocorrelation of the residuals
pacf(residuals(mod1)) # indicates AR1 at lag 10

#accounting for temporal autocorrelation (note basis function with adaptative splines are not possible)
mod1.car <- gamm(diatPrC ~ s(Age, k=40) + s(Ti) + s(pollenPrC) + s(agropastPrC),
                 data = interpolatedData, method = "REML", 
                 weights = elapsedTime_diat/ mean(elapsedTime_diat),
                 family = gaussian(link="identity"),
                 correlation = corCAR1(form = ~ Age),
                 knots = list(negAge=quantile(interpolatedData$Age, seq(0,1, length=10)))) #places notes at the deciles of sample ages

# Check model fit
pacf(residuals(mod1.car$lme))
plot(mod1.car$gam, page=1, scale = 0)
summary(mod1.car$gam)
gam.check(mod1.car$gam)


#Compare different model fits using AIC
AIC_table <- AIC(mod1, mod1.car$lme)%>%
  rownames_to_column(var= "Model")%>%
  mutate(data_source = rep(c("diatom_data")))%>%
  group_by(data_source)%>%
  mutate(deltaAIC = AIC - min(AIC))%>%
  ungroup()%>%
  dplyr::select(-data_source)%>%
  mutate_at(.vars = vars(df,AIC, deltaAIC), 
            .funs = funs(round,.args = list(digits=0)))

AIC_table


# Predict GAM
#Check NAs and remove rows
row.has.na <- apply(interpolatedData, 1, function(x){any(is.na(x))})
sum(row.has.na)
interpolatedData <- interpolatedData[!row.has.na,]

## Predict
predGam <- cbind(interpolatedData, 
                 data.frame(predict.gam(mod1, interpolatedData, 
                                        type = "terms" , se.fit = TRUE)))

#Select covariates to plot (those having significant effects on the mean values of diatom PrC)
var <- predGam$fit.s.agropastPrC.
se.var <- predGam$se.fit.s.agropastPrC.

# Plot temporal contribution of covariates on response 
predGamPlt <- ggplot(predGam, aes(x = Age, y = var)) +
  geom_line() +
  geom_ribbon(aes(ymin = var + (2 * se.var), ymax = var - (2 * se.var), alpha = 0.1)) +
  #scale_x_reverse() +
  labs(y = "Diatom GAM PrC", x = "Cal yr BP", title = "")+
  theme(legend.position = "none")+
  geom_hline(yintercept=0, linetype="dashed")+
  theme_bw()+ theme(legend.position = "none")+
  ggtitle(expression("Agropastoralism" %->% "Diatoms"))

predGamPlt

# Save plot
ggsave("outputs/diatomPrC_GAM_covariates.png",
       plot = predGamPlt ,
       width = 10,
       height=8,
       units="in",
       dpi = 400)


```

# Step 5: Hierarchical Generalized Additive Models (HGAM) 
In this step we model, first, diatom and pollen assemblages using HGAM and then calculate average rate of change at assemblage-level to assess the timing in the start of ecological change and test comparable rates of change between aquatic and terrestrial ecosystems.


By obtaining derivative-based values of the aquatic (diatom) and terrestrial (agropastoralism, pollen, and XRF) time series we obtain regular time intervals at 24-year resolution (median between diatom and pollen records) and hence comparable for the synchronous/asynchronous model. 

```{r HGAM, message=FALSE, error=FALSE, results="hide", size="small"}

```



Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
